---
title: "Voter Data Analysis"
author: "Hamza Khan"
date: "2023-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Importing libraries
```{r warning=FALSE}
library(tidyverse)
library(ggchicklet) # for pretty bar graphs
```

Setting working directory and reading file
```{r warning=FALSE}
setwd("C:/Users/Hamza/OneDrive - University of North Carolina at Chapel Hill/Documents/SP2023/POLI 281/Final")
CES20 <- read_csv("CES20.csv")
```

Keeping needed columns and renaming some of them
```{r warning=FALSE}
CES <- select(CES20,
              faminc_new,
              CC20_402a,
              CC20_433b,
              race,
              educ,
              gender)

colnames(CES)[1] <- "income"
colnames(CES)[2] <- "no_vote"
colnames(CES)[3] <- "partisanship"
```

Filtering income and converting income from being continous variable to a categorical variable
```{r warning=FALSE}
CES_income <- filter(CES, income != 97)
CES$inc_range <- NA
for(i in 1:nrow(CES)){
  if (is.na(CES$income[i])){
    CES$inc_range[i] = "NA"
  }
  else if (CES$income[i] == 1){
    CES$inc_range[i] = "Less than $10,000"
  }
  else if (CES$income[i] == 2){
    CES$inc_range[i] = "$10,000 - $19,999"
  }
  else if (CES$income[i] == 3){
    CES$inc_range[i] = "$20,000 - $29,999"
  }
  else if (CES$income[i] == 4){
    CES$inc_range[i] = "$30,000 - $39,999"
  }
  else if (CES$income[i] == 5){
    CES$inc_range[i] = "$40,000 - $49,999"
  }
  else if (CES$income[i] == 6){
    CES$inc_range[i] = "$50,000 - $59,999"
  }
  else if (CES$income[i] == 7){
    CES$inc_range[i] = "$60,000 - $69,999"
  }
  else if (CES$income[i] == 8){
    CES$inc_range[i] = "$70,000 - $79,999"
  }
  else if (CES$income[i] == 9){
    CES$inc_range[i] = "$80,000 - $99,999"
  }
  else if (CES$income[i] == 10){
    CES$inc_range[i] = "$100,000 - $119,999"
  }
  else if (CES$income[i] == 11){
    CES$inc_range[i] = "$120,000 - $149,999"
  }
  else if (CES$income[i] == 12){
    CES$inc_range[i] = "$150,000 - $199,999"
  }
  else if (CES$income[i] == 13){
    CES$inc_range[i] = "$200,000 - $249,999"
  }
  else if (CES$income[i] == 14){
    CES$inc_range[i] = "$250,000 - $349,999"
  }
  else if (CES$income[i] == 15){
    CES$inc_range[i] = "$350,000 - $499,999"
  }
  else if (CES$income[i] == 16){
    CES$inc_range[i] = "$500,000 or more"
  }
  else if (CES$income[i] == 97){
    CES$inc_range[i] = "Prefer not to say"
  }
}

```

Factorizing inc_range and grouping it 
```{r warning=FALSE}
CES$inc_range <- factor(CES$inc_range, levels=c(
  "Less than $10,000", "$10,000 - $19,999", "$20,000 - $29,999", "$30,000 - $39,999", "$40,000 - $49,999", "$50,000 - $59,999",
  "$60,000 - $69,999", "$70,000 - $79,999", "$80,000 - $99,999", "$100,000 - $119,999", "$120,000 - $149,999", "$150,000 - $199,999",
  "$200,000 - $249,999", "$250,000 - $349,999", "$350,000 - $499,999", "$500,000 or more", "Prefer not to say"))

CES_income_grouped = CES %>%
  group_by(inc_range) %>%
  summarize(inc_count = n())
```


```{r warning=FALSE}
ggplot(data = CES_income_grouped, aes(inc_range, inc_count), !is.na(income)) +
  # geom_bar() +
  geom_chicklet() +
  xlab("Income Level") +
  ylab("Count") +
  theme_bw() +
  ggtitle("Distribution of Income") +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_flip()
# ggsave("income_dist.jpg")
```

Filtering no_vote, recoding values and summarizing
```{r warning=FALSE}
CES_no_vote <- filter(CES, no_vote != 55)

for(i in 1:nrow(CES_no_vote)) {
  if (is.na(CES_no_vote$no_vote[i])){
    CES_no_vote$CES_no_vote[i] = "NA"
  }
  else if (CES_no_vote$no_vote[i] == 1){
    CES_no_vote$no_vote[i] = "Forgot"
  }
  else if (CES_no_vote$no_vote[i] == 2){
    CES_no_vote$no_vote[i] = "Not interested"
  }
  else if (CES_no_vote$no_vote[i] == 3){
    CES_no_vote$no_vote[i] = "Too busy"
  }
  else if (CES_no_vote$no_vote[i] == 4){
    CES_no_vote$no_vote[i] = "Did not like candidates"
  }
  else if (CES_no_vote$no_vote[i] == 5){
    CES_no_vote$no_vote[i] = "Not Registered"
  }
  else if (CES_no_vote$no_vote[i] == 6){
    CES_no_vote$no_vote[i] = "Wrong form of ID"
  }
  else if (CES_no_vote$no_vote[i] == 7){
    CES_no_vote$no_vote[i] = "Out of town"
  }
  else if (CES_no_vote$no_vote[i] == 8){
    CES_no_vote$no_vote[i] = "Sick or disabled"
  }
  else if (CES_no_vote$no_vote[i] == 9){
    CES_no_vote$no_vote[i] = "Transportation"
  }
  else if (CES_no_vote$no_vote[i] == 10){
    CES_no_vote$no_vote[i] = "Bad weather"
  }
  else if (CES_no_vote$no_vote[i] == 11){
    CES_no_vote$no_vote[i] = "Line too long"
  }
  else if (CES_no_vote$no_vote[i] == 12){
    CES_no_vote$no_vote[i] = "Not allowed to vote"
  }
  else if (CES_no_vote$no_vote[i] == 13){
    CES_no_vote$no_vote[i] = "Didn't receive absentee ballot"
  }
  else if (CES_no_vote$no_vote[i] == 14){
    CES_no_vote$no_vote[i] = "Unsure of location"
  }
  else if (CES_no_vote$no_vote[i] == 15){
    CES_no_vote$no_vote[i] = "Lack of info on candidates"
  }
  else if (CES_no_vote$no_vote[i] == 16){
    CES_no_vote$no_vote[i] = "Fear of COVID Exposure"
  }
  else if (CES_no_vote$no_vote[i] == 17){
    CES_no_vote$no_vote[i] = "Other"
  }
  else if (CES_no_vote$no_vote[i] == 55){
    CES_no_vote$no_vote[i] = "Don't know"
  } 
}

CES_no_vote_grouped = CES_no_vote %>%
  group_by(no_vote) %>%
  summarize(no_vote_count = n()) %>% 
  filter(no_vote != "NA")
```


```{r warning=FALSE}
ggplot(data = CES_no_vote_grouped, aes(reorder(no_vote, no_vote_count), no_vote_count), !is.na(no_vote)) +
  # geom_bar() +
  geom_chicklet() +
  xlab("Reason for Not Voting") +
  ylab("Count") +
  coord_flip() +
  # theme_bw() +
  ggtitle("Distribution of reasons for not voting") +
  theme(plot.title = element_text(hjust = 0.5))
# ggsave("dist_no_vote.jpg")
```

Creating a column named income_barrier as a binary variable based off of no_vote
```{r warning=FALSE}
CES$income_barrier <- NA
for(i in 1:nrow(CES)) {
  if (CES$no_vote[i] %in% c("Too busy", "Not Registered", "Wrong form of ID", "Sick or disabled", "Transportation", "Line too long")) {
    CES$income_barrier[i] = 1
  } else {
    CES$income_barrier[i] = 0
  }
}
```

Grouping by inc_range
```{r warning=FALSE}
income_by_reason <- CES %>%
  filter(inc_range != "Prefer not to say") %>%
  group_by(inc_range) %>%
  summarise(avg_barrier = mean(income_barrier))

new_inc_by_res = CES %>%
  filter(no_vote %in% c(
    "Too busy", "Not Registered", "Wrong form of ID", "Sick or disabled", "Transportation", "Line too long")) %>%
  filter(inc_range != "Prefer not to say") %>%
  group_by(inc_range) 
```


```{r warning=FALSE}
ggplot(data = income_by_reason, aes(x = inc_range, y = avg_barrier), !is.na(income)) +
  # geom_col()+
  geom_chicklet() +
  xlab("Level of Income") +
  ylab("Average Voting Barrier") +
  ggtitle("Voting Barriers by Income Level") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_flip()
# ggsave("avg_barrier_inc_level.jpg")
```


```{r warning=FALSE}
ggplot(data = new_inc_by_res, aes(x = inc_range, fill = no_vote), !is.na(income)) +
  geom_bar() +
  # geom_chicklet() +
  xlab("Level of Income") +
  ylab("Average Voting Barrier") +
  theme_bw() +
  ggtitle("Voting Barriers by Income Level") +  
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_flip() +
  labs(fill = "Barrier")
# ggsave("barrier_inc_level.jpg")
```

Initial linear model
```{r warning=FALSE}
fit1 <- lm(CES$income_barrier ~ CES$income)
summary(fit1)
nobs(fit1)
```

Prep work for confounders. Grouping
```{r warning=FALSE}
CES$white <- ifelse(CES$race == 1, 1,0)
CES$educated <- ifelse(CES$educ >= 5, 1,0)
CES$male <- ifelse(CES$gender == 1, 1,0)
CES$democrat <- ifelse(CES$partisanship == 1, 1,0)

cor(CES$income, CES$white, "complete.obs")
cor(CES$income, CES$educated, "complete.obs")
cor(CES$income, CES$male, "complete.obs")
cor(CES$income, CES$democrat, "complete.obs")

# cor(CES$no_vote, CES$white, "complete.obs")
# cor(CES$no_vote, CES$educated, "complete.obs")
# cor(CES$no_vote, CES$male, "complete.obs")
# cor(CES$no_vote, CES$income, "complete.obs")
# cor(CES$no_vote, CES$democrat, "complete.obs")


inc_inc_bar = CES %>%
  group_by(race, income_barrier) %>%
  summarize(counta = count(CES, income_barrier))

```

More linear models
```{r warning=FALSE}
fit1 <- lm(CES$income_barrier ~ CES$income)
summary(fit1)
nobs(fit1)

fit2 <- lm(CES$income_barrier ~ CES$income + CES$educated)
summary(fit2)
nobs(fit2)

fit3 <- lm(CES$income_barrier ~ CES$income + CES$educated + CES$male + CES$white)
summary(fit3)
nobs(fit3)
```

